---
title: "Template for practicals"
author: "The WANT team"
date: "24/08/2018"
output:
  html_document:
    css: want.css
---

# Learning goals
* Understand the concept of local sensitivity analysis
* Understand the limitations of local sensitivity analysis 

# Introduction
In this assignment, we will manually work out the local sensitivity at several places for several functions. The idea is that you can do these calculations by hand. 

# Assignment 1
We have a linear function $y = ax + b$, with parameters $a$ and $b$. 

<span class="question">
Plot the function $y$ for the domain $x = [0, 10]$ in the code chunk below. Base values for $a$ and $b$ are 3 and 7, respectively. Which of the two parameters, $a$ or $b$ do you expect to be of more influence on y? 
</span>

```{r}
# Type your code here.
```

<span class="student_answer">
Fill in your answer...
</span>

<span class="answer">
```{r}
x = seq(0,10,1)
a = 3
b = 7
y = a*x + b
plot(x,y)
```
This is a linear function, where $a$ is the slope of the function and $b$ the intersect. Which one is most sensitive is difficult to determine at this stage. 
</span>

<span class="question">
Determine the sensitivity of y for $a$ and $b$ with the given base-values. You can do this manually, or use R as calculator.  
</span>


```{r}
# Type your code here.
```


<span class="answer">
```{r}
y_base = y
# sensitivity of a
a_


```
This is a linear function, where $a$ is the slope of the function and $b$ the intersect. Which one is most sensitive is difficult to determine at this stage. 
</span>











# Question 2
In this question, you have to make your first plot.

<span class="question">
Plot the function $y = \cos(x)$ for the domain $x = [0, 10]$ in the code chunk below.
</span>

```{r}
# Type your code here.
```

```{r, eval=FALSE}
# This code will not execute
a = c(1,2,3)
```

<span class="answer">
This will only appear in the answers.
```{r}
x = seq(0, 10, 0.1)
y = cos(x)
plot(x,y)
```
</span>

This is the end.
