---
title: "Template for a 1D model"
output: 
  html_document:
    css: want.css
    toc: yes
---

This is a script template to set up a flowmodel in R in combination with package FVFE1D, used in the course Water and Air flow Numerical Techniques.                    R.J. Ronda, P. Torfs and G. Bier - dec. 2015, Wageningen           

# Starting up

## Clear the working directory 
```{r eval=FALSE}
rm(list=ls())
```

## Load the flow modelling package 
```{r eval=FALSE}
library (FVFE1D)
```

# Mathematical model setup

## Define the flow domain of the model
Fill out min and max to determine the length of the model domain, for example `c(0,100)`. c stands for 'concatenate' and results here in a simple array with two values [1]=0 and [2]=100.

```{r eval=FALSE}
domain = c()
```

## Define the system flux function/ internal flux function
Here you implement the expression for the flux of water, energy or something else between the nodes or volumes in the model. To calculate the internal flow the model needs the specific location (x), the state at (x) and the gradient of that state at (x).   
The `return()` argument holds the actual flux-equation, so what is on the right hand side of the flux equation. 

```{r eval=FALSE}
intflux = function(x,state,gradstate)
{  
  Q = ???
  return(Q)
}
```

## Construct the flow model
Two arguments are required:   
1) the domain  
2) the system flux function 

```{r eval=FALSE}
oneDmodel = newFLOW1D(domain=domain,systemfluxfunction=intflux,name="example model")
```

## Add external fluxes
In most cases there is not only an internal flux but also external fluxes.   
Two generic flavours are available:   
1) `add.pointflux(model, at, value, name = "")`   
with arguments:   
`model` : the one dimensional model at hand   
`at`    : the location in the domain where to add the point flux. For example at 50.7 m.   
`name`  : optional; to give the point flux a specific name. This name will turn up in the model outcomes.   
2) `add.spatialflux(model, toadd, name = "")`   
with arguments:   
`model` : the one dimensional model at hand   
`toadd` : either a function of space (x) and state (s) returning  the intensity of the spatial flux at that point; or a constant giving the constant flux at all spatial points for all state values.   
`name`  : optional; to give the point flux a specific name   

```{r eval=FALSE}
flux=0.1
add.spatialflux(oneDmodel,flux,name='constant inflow')
```

## Set the boundary conditions

Now also boundary conditions need to be set.       
Three types of Boundary Conditions (BC) can be applied:      
* **Dirichlet BC** -> a function to set a prescribed state.   
  usage: `set.BC.fixedstate(model,where, value)`   
* **Neumann BC** -> a function to set a prescribed prescribed flux.     
  usage: `set.BC.fixedflux(model,where, value)`     
* **Cauchy/Robin/Fourier BC** -> a function to formulate a state dependent flux.     
  usage: `set.BC.stateflux(model,where,value)`   

These functions have three arguments:   
* `model` = the current model   
* `where` = a string to set the side of the BC: 'l' or 'left, 'r' or 'right'    
* `value` = the value (or function) to be assigned to this BC.   
Omitting a boundary condition at the boundary will be interpreted as a no-flow boundary condition.     
In case of vertical flow the 'left' boundary condition is actually the BOTTOM boundary condition. The 'right' boundary condition is interpreted as the TOP boundary condition.

```{r eval=FALSE}
BCleft=???
BCright=???
set.BC.fixedstate(oneDmodel,'left',BCleft) 
set.BC.fixedstate(oneDmodel,'right',BCright)
```

## Model name
You can change the name of your model with:
```{r eval=FALSE}
set.name(model=oneDmodel,newname='example model')
```

The model is now constructed; internal and external fluxes are defined; and boundary conditions are set.

# Numerical model setup 

## Discretisation
Before the model can be solved, it first needs to be discretised.
Discretisation involves two steps:   
1) setting the nodal distribution in the model domain.      
Nodes can be created with the sequence function `seq`, using the model domain and the required nodal distance.

```{r eval=FALSE}
nodes = seq(from=???,to=???,by=???)
```

2) setting the discretisation method.      
This is done by: `set.discretisation (model, nodes, method)`   
with:   
`model` : the model for which the states (at the nodes) are solved   
`nodes` : the nodes in the domain at which discrete approximations of the state will be calculated   
`method`: choice of discretisation method. The following choices are implemented:    
* `"FV"` - finite volumes, which we will use most   
* `"FE"`,`"FElinear"`
* `"FEquadratic"`,`"FEcubic"`,`"FEquartic"`
 
```{r eval=FALSE}
set.discretisation(oneDmodel, nodes, method= "FV")
```

## Solving the model 
The model needs to be solved now in such a way that either the change in state (stationary) or the misfit in the fluxes (transient) are minimised until a certain criterion.
The function `solve.step(model)` does *one* adjustment of all states in the domain. The function `solve.steps(model)` solves the model by taking max. 500 updating steps, or until the root mean squared misfit is less than 0.0005.   
The `verboselevel`argument determines if the progress is printed to the screen.    
The `control` variable contains two results:   
* 'RMSM', root mean square misfit and    
* 'MAM', max absolute misfit   
```{r eval=FALSE}
control = solve.steps(oneDmodel, verboselevel=1)
```

# Postprocessing
After the model is solved for the states, one can evaluate the results. An enormous palette of plots is available. See R related URL's for details.
Here we just give two examples:

```{r eval=FALSE}
plot(oneDmodel)
```
This plots the state values on the y-axis and the cell centers on the x-axis.

```{r eval=FALSE}
plot(oneDmodel,stateplot=TRUE,fluxplot=TRUE)
```
This plots the state values on the y-axis and the cell centers on the x-axis, and adds a plot giving the fluxes.

To have a look at flux and state values in the model one can make use of the data frame functions included in the package.
Five data frames are predefined and can directly be used for further analyses:   
* `dataframe.balance`       : overview of the balance    
* `dataframe.boundaries`    : overview of the boundary fluxes   
* `dataframe.externalfluxes`: overview of the external fluxes   
* `dataframe.internalfluxes`: overview of the internal fluxes (between neighbouring nodes)   
* `dataframe.states`        : overview of the states at the nodes in the model   
