---
title: "Template for practicals"
author: "The WANT team"
date: "24/08/2018"
output:
  html_document:
    css: want.css
    toc: yes
---
<!-- include the hint.js -->
<script src="hints.js"></script>

# Learning goals
* Goal 1
* Goal 2

# Introduction
In this tutorial, you will learn a lot.

# Question 1
In this question, you have to make your first plot.

<span class="question">
Plot the function $y = \sin(x)$ for the domain $x = [0, 10]$ in the code chunk below.
</span>

```{r}
# Type your code here.
```

<!-- Here a hint is implemented. Argument to showHint is the QUESTION_NUMBER as documented in the hints.js file.
The position the hint is displayed where a paragraph text with id = Q + QUESTION_NUMBER is entered. -->
<button type="button" onclick="showHint(1)">Show Hint</button>
<p id="Q1"> </p>

<span class="answer">
This will only appear in the answers.
```{r}
x = seq(0, 10, 0.1)
y = sin(x)
plot(x,y)
```
</span>

<span class="student_answer">
Fill in your answer...
</span>

<span class="answer">
This is the correct anwer.
</span>


# Question 2
In this question, you have to make your first plot.

<span class="question">
Plot the function $y = \cos(x)$ for the domain $x = [0, 10]$ in the code chunk below.
</span>

```{r}
# Type your code here.
```

<button type="button" onclick="showHint(2)">Show Hint</button>
<p id="Q2"> </p>

```{r, eval=FALSE}
# This code will not execute
a = c(1,2,3)
```

<span class="answer">
This will only appear in the answers.
```{r}
x = seq(0, 10, 0.1)
y = cos(x)
plot(x,y)
```
</span>

This is the end.
