---
title: "Finite differences 2 Jasper Question"
author: "The WANT team"
date: "24/09/2018"
output:
  html_document:
    css: ../../want.css
    toc: yes
    toc_depth: 2
---
<!-- include the hint.js -->
<script src="../../hints.js"></script>


For the first example with the three differencing schemes we will examine the simplest wave equation possible, a sinus-function ($s(x) = \sin(x)$) with it's derivative that is the opposite cosine wave ($\frac{ds}{dx}=\cos(x)$). The first chunk is fully preprogrammed, it is exactly the same as the approach used for time integration. Easiest way to understand it is to work from left to right (instead of from the beginning to the end as we did with the time dimension). An forward integration scheme is used, meaning that information of one point is used to calculate the point directly to the right of it. Still the same stuff as in the previous practical but it's maybe a bit less intuitive. The following information about the system is available:

$$
\begin{matrix}
s(x) &=& \sin(x)\\
\dfrac{ds}{dx} &=& \cos(x)\\
\dfrac{s(x+\Delta x) - s(x)}{\Delta x} &=& \dfrac{ds}{dx}
\end{matrix}
$$
Leading to:
$$
\begin{matrix}
\dfrac{s(x+\Delta x) - s(x)}{\Delta x} &=& cos(x)\\
s(x+\Delta x) &=& s(x) + \Delta x cos(x)
\end{matrix}
$$

To implement this equation a boundary state should be set. In this case 0 is a logical choice. This is equivalent to setting an initial state for time integration. For ease if visualisation the right side coordinates (end.time in previous) is set to $2\pi$ and dx to $\frac{\pi}{6}$.  

<span class="comment"> This would be enough information to actually implement this chunk. Do we want to let them do that or does it cost too much time? </span>

```{r simple_forward}
# Initial parameter values
left          = 0     # usually this variable is set to 0
right         = 2*pi  # end time of the simulation (50)
dx            = pi/6  # delta t; time discretisation (0.25)
left.state    = 0     # state of the system at the beginning of the simulation

# Simulation initialisation
space         = left
result.state  = c(left.state)
result.space  = c(left)
current.state = left.state

sys.fun = function(space,state){
  return(state + dx*cos(space))
}

# Simulation
while(space < right)
{
  current.state = sys.fun(space,current.state)
  result.state  = c(result.state,current.state)
  space         = space + dx
  result.space  = c(result.space,space)
}

# Plotting
plot(result.space,result.state,type='o', xlab="x (m)",ylab="state (m)", ylim = c(-1.5,1.5), col="blue")
space.sequence = seq(left,right,by=0.05)
analyt.state=sin(space.sequence)
lines(space.sequence,analyt.state,col='grey',lwd=2)
title(main='Numerical and analytical solution compared')
grid()
```

As introduced in the Part 2 of this practical, this calculation can be done in a quicker way with exactly the same result by setting up a matrix equation. The following script is the same system as the one described above, but implemented by using matrix calculus. The first part sets up the initial parameters. V is the right hand side of the equation, M is the system matrix. Then the matrix and right hand side vector are populated. The equation is solved by running the solve(M,V) command and afterwards the result is plotted. Populating the matrix works as follows (as an example):

* First set up the nodes of the matrix: 5 (in the chunk more nodes are used)
* Defining the equations (square brackets don't signify the location of the coordinate but the index in the matrix):

$$
\begin{matrix}
\dfrac{s[2]}{\Delta x}  &=& \dfrac{s[1]}{\Delta x}  &+& cos(x)\\
\dfrac{s[3]}{\Delta x}  &=& \dfrac{s[2]}{\Delta x}  &+& cos(x)\\
\dfrac{s[4]}{\Delta x}  &=& \dfrac{s[3]}{\Delta x}  &+& cos(x)\\
\dfrac{s[5]}{\Delta x}  &=& \dfrac{s[4]}{\Delta x}  &+& cos(x)
\end{matrix}
$$

* Translating them to matrix, vector logic (with the vector entries on the right-hand side:

$$
\begin{matrix}
\dfrac{s[2]}{\Delta x}  &-& \dfrac{s[1]}{\Delta x}  &=& cos(x)\\
\dfrac{s[3]}{\Delta x}  &-& \dfrac{s[2]}{\Delta x}  &=& cos(x)\\
\dfrac{s[4]}{\Delta x}  &-& \dfrac{s[3]}{\Delta x}  &=& cos(x)\\
\dfrac{s[5]}{\Delta x}  &-& \dfrac{s[4]}{\Delta x}  &=& cos(x)\\
\end{matrix}
$$
This system is underfined if left like this (4 equations with 5 variables). A boundary condition is needed to solve this problem. The easiest boundary condition is the Dirichlet boundary condition where you prescribe one of the states as was done in the chunk above and also in the previous practical (a initial state). This is done with the following equation:

$$ s[1] = 0$$

<span class="question"> 
3.1 Insert the formulas stated above in the following matrix equation
$$ 
M * s = V:
$$
</span>

<span class="answer">
$$
\begin{bmatrix}
&1&.&.&.&.\\
&\frac{-1}{\Delta x}&\frac{1}{\Delta x}&.&.&.\\
&.&\frac{-1}{\Delta x}&\frac{1}{\Delta x}&.&.\\
&.&.&\frac{-1}{\Delta x}&\frac{1}{\Delta x}&.\\
&.&.&.&\frac{-1}{\Delta x}&\frac{1}{\Delta x}
\end{bmatrix}
* 
\begin{bmatrix}
s[1]\\ 
s[2]\\ 
s[3]\\ 
s[4]\\ 
s[5]
\end{bmatrix}
= 
\begin{bmatrix}
0\\
cos(x)\\ 
cos(x)\\ 
cos(x)\\ 
cos(x)
\end{bmatrix}
$$
</span>

To solve the states, matrix algebra is used:
$$
\begin{matrix}
M  s &=& V\\
M^{-1}M s &=& M^{-1}V\\
I s &=& M^{-1}V\\
s &=& M^{-1}V
\end{matrix}
$$
*where I is the identity matrix*

<span class="question">
3.2 A boundary value is needed to solve the equations. Explain the reason using matrix algebra
</span>

<span class="answer">
Determinant is 0, which means the matrix is ininvertable. This leads to an infinite amount of solutions for this equation.
</span>

<span class="question">

<span class="comment"> maybe the hard track can do this entirely by themselves </span>

3.3 complete the following chunk by constructing the matrix and inserting boundary values as shown above in mathematical notation. If everything went well you will end up with exactly the same plot as the one generated by chunk *simple forward*.
</span>

<span class="student_answer">
```{r matrix_forward, eval=FALSE}
rm(list=ls()) 

# setting up initial parameters
dx    = pi/6
left  = 0                      # left domain
right = 2*pi                   # right domain
xcoor = seq(left,right, by=dx) # builds a list of all coordinates

bl    = 0                      # left boundary condition
V     = c()                    # build an empty vector for the solutions
M     = matrix(0,nrow=length(xcoor),ncol=length(xcoor)) # an empty matrix for the differencing scheme

### FILL IN THE MATRIX AND VECTOR HERE

# solve matrix equation
s = solve(M,V)

# plotting
plot(xcoor,s,ylab="state (m)",xlab="x (m) ",type='o',col='blue', ylim=c(-1.5,1.5)) # numerical approximation
x_sol = seq(left,right,by=0.05)
lines(x_sol,sin(x_sol),type='l', col="grey",lwd=2)
title(main='Forward differencing scheme with matrices')
grid()
```
</span>

<span class="answer">
```{r matrix_forward_answer_left_boundary}
rm(list=ls()) 

# setting up initial parameters
dx    = pi/6
left  = 0
right = 2*pi
xcoor = seq(left,right, by=dx) # builds a list of all coordinates
N     = length(xcoor)

bl    = 0                      # left boundary condition
V     = c()                    # build an empty vector for the solutions
M     = matrix(0,nrow=length(xcoor),ncol=length(xcoor)) # an empty matrix for the differencing scheme

# populating vector and matrix
for(j in 2:(length(xcoor)))
{
  M[j,j]   = 1/dx
  M[j,j-1] = -1/dx
  V[j]     = cos(xcoor[j-1]) # it belongs to the j-1 step
}

# left boundary condtion
M[1,1]     = 1
V[1]       = bl

# solve matrix equation
s = solve(M,V)

# plotting
plot(xcoor,s,ylab="state (m)",xlab="x (m) ",type='o',col='blue', ylim=c(-1.5,1.5)) # numerical approximation
x_sol = seq(left,right,by=0.05)
lines(x_sol,sin(x_sol),type='l', col="grey",lwd=2)
title(main='Forward differencing scheme with matrices')
grid()
```

<span class="question">
3.4 Use the same system above but instead of using a left boundary condition of $s[1] = 0$, use $s[5] = 0$. Implement it using the previous chunk and show that this will result in exactly the same plot (because $\sin(0) = \sin(2\pi) = 0$). TIP! If needed write down the matrix by hand as done in the previous exercise.
</span>

<span class="answer">
```{r matrix_forward_answer}
rm(list=ls()) 

# setting up initial parameters
dx    = pi/6
left  = 0
right = 2*pi
xcoor = seq(left,right, by=dx) # builds a list of all coordinates
N     = length(xcoor)

br    = 0                      # left boundary condition
V     = c()                    # build an empty vector for the solutions
M     = matrix(0, nrow=length(xcoor), ncol=length(xcoor)) # an empty matrix for the differencing scheme

# populating vector and matrix
for(j in 1:(length(xcoor)-1))
{
  M[j,j]   = -1/dx
  M[j,j+1] = 1/dx
  V[j]     = cos(xcoor[j]) # it belongs to the j'th step
}

# left boundary condtion
M[N,N]     = 1
V[N]       = br

# solve matrix equation
s = solve(M,V)

# plotting
plot(xcoor,s,ylab="state (m)",xlab="x (m) ",type='o',col='blue', ylim=c(-1.5,1.5)) # numerical approximation
x_sol = seq(left,right,by=0.05)
lines(x_sol,sin(x_sol),type='l', col="grey",lwd=2)
title(main='Forward differencing scheme with matrices')
grid()
```
</span>

<span class="question">
3.5 Copy the previous chunk and adapt it to the backward differencing scheme (still explicit integration method)
</span>

<span class="answer">
```{r backward}
rm(list=ls()) 

# setting up initial parameters
dx    = pi/6
left  = 0
right = 2*pi
xcoor = seq(left,right, by=dx) # builds a list of all coordinates

bl    = 0                      # left boundary condition
V     = c()                    # build an empty vector for the solutions
M     = matrix(0,nrow=length(xcoor),ncol=length(xcoor)) # an empty matrix for the differencing scheme

# populating vector and matrix
for(j in 2:(length(xcoor)))
{
  M[j,j]   = 1
  M[j,j-1] = -1
  V[j]     = dx*cos(xcoor[j])
}

# left boundary condtion
M[1,1]     = 1
M[1,2]     = 0
V[1]       = bl

# solve matrix equation
s = solve(M,V)

# plotting
plot(xcoor,s,ylab="state (m)",xlab="x (m) ",type='o',col='blue', ylim=c(-1.5,1.5)) # numerical approximation
x_sol = seq(left,right,by=0.05)
lines(x_sol,sin(x_sol),type='l', col="grey",lwd=2)
title(main='Backward differencing scheme with matrices')
grid()
```
</class>

<span class="comment"> I remove central scheme here because it's slightly tricky but I come back to this in Part 4 </span>
