---
title: "Finite differences 1 - extended questions"
author: "The WANT team"
date: "24/09/2018"
output:
  html_document:
    css: ../../want.css
---

## Exercise 1
<span class="question">
Add the analytical solution of equation to chunk3.
</span>

<span class = "answer">
```{r eval=FALSE}
time.sequence = seq(begin.time,end.time,by=0.1)
analyt.state=init.state * exp(-time.sequence*alpha)
lines(time.sequence,analyt.state,col='grey',lwd=2)
```
</span>

## Exercise 2
<span class="question">
Set up an equation to calculate the error. Use equations 3 and 8.
</span>

<span class = "answer">
Starting from equation 8:
$$s(t + \Delta t) = s(t) + \frac{\Delta t}{1!} \frac{ds}{dt} \bigg\rvert_{t} + \frac{\Delta t^{2}}{2!} \frac{d^{2}s}{dt^{2}} \bigg\rvert_{t} + \frac{\Delta t^{3}}{3!} \frac{d^{3}s}{dt^{3}} \bigg\rvert_{t} + \frac{\Delta t^{4}}{4!} \frac{d^{4}s}{dt^{4}} \bigg\rvert_{t} + \ ...     \tag{10}$$
Rearranging this equation:

$$\frac{s(t + \Delta t) - s(t)}{\Delta t} =  \frac{ds}{dt} \bigg\rvert_{t} + \frac{\Delta t}{2!} \frac{d^{2}s}{dt^{2}} \bigg\rvert_{t} + \frac{\Delta t^{2}}{3!} \frac{d^{3}s}{dt^{3}} \bigg\rvert_{t} + \frac{\Delta t^{3}}{4!} \frac{d^{4}s}{dt^{4}} \bigg\rvert_{t} + \ ... \tag{11}$$
Comparing this with the implementation in exercise 1 leads to the quantification of the error:

$$\frac{s(t + \Delta t) - s(t)}{\Delta t} =  \frac{ds}{dt} \bigg\rvert_{t} + error    \tag{12}$$

$$ error = \frac{\Delta t}{2!} \frac{d^{2}s}{dt^{2}} \bigg\rvert_{t} + \frac{\Delta t^{2}}{3!} \frac{d^{3}s}{dt^{3}} \bigg\rvert_{t} + \frac{\Delta t^{3}}{4!} \frac{d^{4}s}{dt^{4}} \bigg\rvert_{t} + \ ... \tag{13}$$
</span>

## Exercise 3
<span class="question">

use the R-functions *expression()*, *D()* (derivative) and *eval()* to implement the first six terms of equation 10 incremently in chunk5 below and plot this result on top of the existing plot. Note down the error every time a term is added.

terms | error 
----- | ----- 
1 |
2 |  
3 |
4 |
5 | 
6 | 

```{r chunk5_question, eval=FALSE}
# Initial parameter values
begin.time = 0     # usually this variable is set to 0
end.time   = 50    # end time of the simulation (50)
dt         = 5     # delta t; time discretisation (0.25)
init.state = 3     # state of the system at the beginning of the simulation (3)
alpha      = 0.25  # decay constant (0.25)

# System function
sys.fun = function(state)
{
  newstate = state*(1-alpha*dt)
  return(newstate)
}

# Simulation initialisation
time = begin.time
result.state= c(init.state)
result.time = c(time)
current.state = init.state

# Simulation
while(time < end.time)
{
  current.state = sys.fun(current.state)
  result.state = c(result.state,current.state)
  time = time + dt
  result.time = c(result.time,time)
}

# Taylor series approximation
f  = ...                 

DT = seq(begin.time,end.time,by=0.1) # list of evaluation times
p  = eval(f)                         # list of results of the implemented funtion

# Plotting
plot(result.time,result.state,type='o', xlab="time",ylab="state")
time.sequence = seq(begin.time,end.time,by=0.1)
analyt.state=init.state * exp(-time.sequence*alpha)
lines(time.sequence,analyt.state,col='blue',lwd=2)
lines(DT,p, lwd=2, col="green")

DT = dt
correct.state = ...     # state at 1 step (dt)
error = ...             # determine the distance between the correct.state and the value calculated with the taylor series 
points(dt,eval(g),col="green",pch=19)
points(dt,correct.state,pch=19)
segments(dt,correct.state,dt,eval(g),col="red",lwd=2)
title(main=paste('Numerical and analytical solution compared. Error: ',error))
      
grid()
```

</span>

<span class="answer">
answer hard question

```{r chunk5_answer}
rm(list=ls())
# Initial parameter values
begin.time = 0     # usually this variable is set to 0
end.time   = 50    # end time of the simulation (50)
dt         = 5     # delta t; time discretisation (0.25)
init.state = 3     # state of the system at the beginning of the simulation (3)
alpha      = 0.25  # decay constant (0.25)

# System function
sys.fun = function(state)
{
  newstate = state*(1-alpha*dt)
  return(newstate)
}

# Simulation initialisation
time = begin.time
result.state= c(init.state)
result.time = c(time)
current.state = init.state

# Simulation
while(time < end.time)
{
  current.state = sys.fun(current.state)
  result.state = c(result.state,current.state)
  time = time + dt
  result.time = c(result.time,time)
}

# Taylor series approximation

f = expression((-0.75 + 0.1875 * DT^(1)/2 - 0.046875*DT^(2)/6) * DT + 3)   
# constants obtained by integration and evaluation (function D() and eval())

DT = seq(begin.time,end.time,by=0.1)  # list of evaluation times
p  = eval(f)                          # list of results of the implemented funtion

# Plotting
plot(result.time,result.state,type='o', xlab="time",ylab="state")
time.sequence = seq(begin.time,end.time,by=0.1)
analyt.state=init.state * exp(-time.sequence*alpha)
lines(time.sequence,analyt.state,col='blue',lwd=2)
lines(DT,p, lwd=2, col="green")

DT = dt
correct.state = init.state*exp(-dt*alpha)
error = round(abs(correct.state-eval(f)),3)
points(dt,eval(f),col="green",pch=19)
points(dt,correct.state,pch=19)
segments(dt,correct.state,dt,eval(f),col="red",lwd=2)
title(main=paste('Numerical and analytical solution compared. Error: ',error))

grid()
```

terms | error 
----- | ----- 
1 | 2.140
2 | 1.609 
3 | 0.242 
4 | 0.063 
5 | 0.013 
6 | 0.002 

The error rapidly decreases every time the next term of the taylor series is added.

</span>

## Exercise 4
<span class="question">
<span class="comment">  Is this question too complex for the hard track? </span>

calculate the following ratio and explain it's meaning (limit $\Delta t$ to 0):

$$ \lim_{\Delta t\to0}\dfrac{Error_{explicit}}{Error_{midpoint}} $$
by making the following assumption ($i$ equals every positive integer):

$$
\dfrac{d^is}{dt^i}\bigg\rvert_{t} = \dfrac{d^is}{dt^i}\bigg\rvert_{t+\Delta t} = 1
$$
</span>

<span class="answer">

Defenition $Error_{midpoint}$

\begin{matrix}

s(t+\frac{1}{2}\Delta t) = s(t) + \frac{1}{2}\Delta t \dfrac{ds}{dt}\bigg\rvert_{t} + \sum_{i=2}^{\infty} \dfrac{(\frac{1}{2}\Delta t)^i}{i!}\dfrac{d^is}{dt^i}\bigg\rvert_{t}\\

\tilde{s} = s(t) + \frac{1}{2}\Delta t \dfrac{ds}{dt}\bigg\rvert_{t} \approx s(t+\frac{1}{2}\Delta t)\\

s(t+\Delta t) = s(t+\frac{1}{2}\Delta t) + \frac{1}{2}\Delta t \dfrac{ds}{dt}\bigg\rvert_{t+\Delta t} + \sum_{i=2}^{\infty} \dfrac{(\frac{1}{2}\Delta t)^i}{i!}\dfrac{d^is}{dt^i}\bigg\rvert_{t+\Delta t}\\

s(t+\Delta t) \approx s(t+\frac{1}{2}\Delta t) + \frac{1}{2}\Delta t \dfrac{ds}{dt}\bigg\rvert_{t + \Delta t}\\

Error_{midpoint} = \sum_{i=2}^{\infty} \dfrac{(\frac{1}{2}\Delta t)^i}{i!} (\dfrac{d^is}{dt^i}\bigg\rvert_{t} + \dfrac{d^is}{dt^i}\bigg\rvert_{t+\Delta t})\\

Error_{midpoint} = \sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{2^{i-1}i!} \dfrac{d^is}{dt^i}\bigg\rvert_{t}

\end{matrix}

Defenition $Error_{explicit}$

\begin{matrix}

s(t+\Delta t) = s(t) + \Delta t \dfrac{ds}{dt}\bigg\rvert_{t} + \sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{i!} \dfrac{d^is}{dt^i}\bigg\rvert_{t} \\

s(t+\Delta t) \approx s(t) + \Delta t \dfrac{ds}{dt}\bigg\rvert_{t}\\

Error_{explicit} = \sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{i!} \dfrac{d^is}{dt^i}\bigg\rvert_{t} 

\end{matrix}

Calculating the ratio:
\begin{align}

Error_{explicit}/Error_{midpoint} &=& \dfrac{\sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{i!} \dfrac{d^is}{dt^i}\bigg\rvert_{t} }{\sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{2^{i-1}i!} \dfrac{d^is}{dt^i}\bigg\rvert_{t}}\\

Error_{explicit}/Error_{midpoint} &=& \dfrac{\sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{i!}}{\sum_{i=2}^{\infty} \dfrac{(\Delta t)^i}{2^{i-1}i!}}\\

Error_{explicit}/Error_{midpoint} &=& \dfrac{-\Delta t + e^{\Delta t} - 1}{-\Delta t + 2e^{\frac{\Delta t}{2}} - 2}

\end{align}

\begin{align}
\lim_{\Delta t\to0}\dfrac{-\Delta t + e^{\Delta t} - 1}{-\Delta t + 2e^{\frac{\Delta t}{2}} - 2} = 2

\end{align}

with a $\Delta t$ approximating 0 the error of the midpoint method is twice as low as the explicit methodology, which makes sense because of the "fake" extra point that is added. The real value of this methodology becomes apparent with increasing $\Delta t$ when the ratio increases rapidly.

</span>





